--- 
title: "Measuring Library Demand: A Novel Data-Driven Library Demand Index and Its Predictors" 
subtitle: "R Code"
author: "Noémi Erika Békésiné Bognár, László Kovács"
date: "August 3, 2026"
output:
  html_document:
    toc: true
    toc_float: true
    number_sections: true
    df_print: paged
    self_contained: true
---

# PREPARATIONS

## Set the working directory

```{r setup, include=FALSE}
knitr::opts_knit$set(root.dir = "C:/tobbvaltozos")

knitr::opts_chunk$set(
  warning = FALSE,
  message = FALSE
)  
```

## Install and load the required packages

```{r}
#install.packages("readxl")
library(readxl)

#install.packages("writexl")
library(writexl)

#install.packages("ggplot2")
library(ggplot2)

#install.packages("GGally")
library(GGally)

#install.packages("psych")
library(psych)

#install.packages("factoextra")
library(factoextra)

#install.packages("lmtest")
library(lmtest)

#install.packages("car")
library(car)

#install.packages("MASS")
library(MASS)

#install.packages("performance")
library(performance) 

#install.packages("dplyr")
library(dplyr)
```

## Import the cleaned public library statistical dataset for 2022

```{r}
elemzes_input_output_cleaned <- read_excel("elemzes_input_output_cleaned.xlsx", sheet = 1)
```

Convert data types

```{r}
elemzes_input_output_cleaned <- data.frame(lapply(elemzes_input_output_cleaned, function(x) {
  if (is.matrix(x) || is.array(x)) as.numeric(x) else x
}))
```

## Dataset structure

### Basic variables

- **ID10 (character)**: library identifier
- **AAA10 (character)**: library name
- **AIF40 (character)**: library type
- **AIF50 (character)**: library type 2
- **AAC10 (character)**: municipality of the reporting library
- **TAB00 (character)**: settlement category
- **HNT (numeric)**: population
- **TAC00 (character)**: county
- **HNT_J (character)**: district
- **osszevont (numeric)**: merged institution 

### Input variables

- **IP1 (numeric, log scale)**: Library funding: level of library expenditure per capita (total expenditure / population)
- **IP2 (numeric, log scale)**: Share of document acquisition in total expenditure (collection development expenditure based on paid invoices / total expenditure)
- **IP3 (numeric, log scale)**: Acquisition quota: collection development expenditure per capita (collection development expenditure based on paid invoices / population)
- **IL1 (numeric, log scale)**: Floor area provision: user space per capita (area of spaces visited and used by users (m²) / population)
- **IL2 (numeric)**: Number of opening hours in an average week
- **IK1 (numeric, log scale)**: Provision of new acquisitions: additions to the collection per capita (number of documents added to the collection in the reference year / population)
- **IK2 (numeric, log scale)**: Document provision: size of the collection per capita (size of the collection as of 31 December of the reference year / population)
- **IK3 (numeric, log scale)**: Proportion of digital documents in new acquisitions ((number of digital documents + e-books) / number of documents added to the collection in the reference year)
- **ISZ1 (numeric)**: Registration fee (Y = 1, N = 0)
- **ISZ2 (numeric)**: Home delivery (Y = 1, N = 0)
- **ISZ3 (numeric)**: Self-service borrowing (Y = 1, N = 0)
- **ISZ4 (numeric)**: Remote service: remote renewal (Y = 1, N = 0)
- **ISZ5 (numeric)**: Remote service: remote payment (Y = 1, N = 0)
- **II1 (numeric, log scale)**: Internet provision: number of internet-connected computers per capita (number of all public, networked, internet-capable computers / population)
- **II2 (numeric)**: Are computers, workstations and mobile devices on average less than 5 years old? (Y = 1, N = 0)
- **IH1 (numeric, log scale)**: Professional staffing: number of professionally employed staff per capita (number of staff employed in professional positions / population)
- **IR1 (numeric, log scale)**: Event provision: number of library events per capita (number of library programmes and training sessions / population)

### Output variables

- **OGY1 (numeric, log scale)**: Popularity of lending: number of loans per capita (number of loans / population)
- **OGY2 (numeric, log scale)**: Popularity of in-house use: number of in-house documents used per capita (number of in-house documents used / population)
- **OGY3 (numeric, log scale)**: Lending activity: number of loans per registered member (number of loans / number of registered users)
- **OA1 (numeric, log scale)**: Proportion of documents borrowed from the lendable collection (number of loans / number of items in the lendable collection)
- **OA2 (numeric, log scale)**: Proportion of documents used in-house from the total collection (number of in-house documents used / number of items in the collection)
- **OK1 (numeric, log scale)**: Popularity of in-person use: number of in-person visits per capita (number of in-person visits / population)
- **OK2 (numeric, log scale)**: Popularity of remote use: number of remote use occasions per capita (number of remote use occasions / population)
- **OKH1 (numeric, log scale)**: Proportion of registered users/readers in the population (number of registered users / population)
- **OE1 (numeric, log scale)**: Popularity of events: number of user participations in library events per capita (number of event participants / population)
- **OE2 (numeric, log scale)**: Attendance per event: number of participants per event (number of event participants / number of events)


## Review the dataset structure

```{r}
glimpse(elemzes_input_output_cleaned)
```

## Summary statistics

```{r}
summary(elemzes_input_output_cleaned)
```

# PRINCIPAL COMPONENT ANALYSIS

## Assess the assumptions

Preliminary tests

- **Bartlett's test of sphericity:**

```{r}
#install.packages("psych")
#library(psych)  

# Output variables
vars_output <- c("OGY1", "OGY2", "OGY3", "OA1", "OA2", "OK1", "OK2", "OKH1", "OE1", "OE2")
pca_output <- elemzes_input_output_cleaned[vars_output]

bartlett_result_output <- cortest.bartlett(cor(pca_output), n = nrow(pca_output))
print(bartlett_result_output)

```
- Kaiser-Meyer-Olkin (KMO) test:

```{r}
kmo_result_output <- KMO(cor(pca_output))

print(kmo_result_output)

```
## Perform principal component analysis

```{r}
fokomp_output <- prcomp(elemzes_input_output_cleaned[,27:36], center = TRUE, scale=TRUE)
str(fokomp_output)
```
```{r}
summary(fokomp_output)
```
Scree plots

```{r}
#library(factoextra)

# Scree plot
fviz_eig(fokomp_output, addlabels = TRUE, ylim = c(0, 100), main = "Scree Plot of Eigenvalues")

```

```{r}
# Basic scree plot with eigenvalues
screeplot(fokomp_output, type = "lines", main = "Scree Plot")
```
Results

```{r}
elemzes_pca_output <- cbind(elemzes_input_output_cleaned, fokomp_output$x[,1:4])
str(elemzes_pca_output)
```

Correlation heatmap
```{r}
#install.packages("dplyr")
#library(dplyr)

Korr_output <- cor(elemzes_pca_output %>%
                   dplyr::select(27:ncol(elemzes_pca_output)) %>% 
                   dplyr::select(-AIF50))

library(corrplot)

corrplot(Korr_output, method="color")
```

## Varimax rotation

```{r}
# # Perform principal component analysis 
fokomp_output_varimax <- prcomp(elemzes_input_output_cleaned[,27:36], center = TRUE, scale = TRUE)

# Apply varimax rotation to the first four principal components
varimax_rotation_varimax <- varimax(fokomp_output_varimax$rotation[,1:4])

# Create a new dataset
varimax_output_df_varimax <- scale(elemzes_input_output_cleaned[,27:36]) %*% varimax_rotation_varimax$loadings

```

Results

```{r}
summary(fokomp_output_varimax)
```
Extract summary statistics into a table

```{r}
summary_prcomp <- summary(fokomp_output_varimax)

standard_deviation <- fokomp_output_varimax$sdev
proportion_of_variance <- summary_prcomp$importance[2, ]
cumulative_proportion <- summary_prcomp$importance[3, ]

prcomp_summary_df <- data.frame(
  Component = seq_along(standard_deviation),
  Standard_Deviation = standard_deviation,
  Proportion_of_Variance = proportion_of_variance,
  Cumulative_Proportion = cumulative_proportion
)

write.csv(prcomp_summary_df, "prcomp_summary.csv", row.names = FALSE)

```

```{r}
# Append the principal component scores to the original dataset
elemzes_pca_output_varimax <- cbind(elemzes_input_output_cleaned, varimax_output_df_varimax)

```

```{r}
# Calculate the correlation matrix (excluding AIF50)
Korr_output_varimax <- cor(elemzes_pca_output_varimax %>%
                           dplyr::select(27:ncol(elemzes_pca_output_varimax)) %>%
                           dplyr::select(-AIF50))

# Visualise the correlation matrix after varimax rotation
library(corrplot)
corrplot(Korr_output_varimax, method = "color")

```

```{r}
print(Korr_output_varimax)
```
## Principal component analysis without OK2 using varimax rotation

```{r}
# Perform principal component analysis without OK2
fokomp_output_varimax_OK2 <- prcomp(elemzes_input_output_cleaned[,c(27:32, 34:36)], center = TRUE, scale = TRUE)

# Apply varimax rotation to the first four principal components
varimax_rotation_varimax_OK2 <- varimax(fokomp_output_varimax_OK2$rotation[,1:4])

# Create a new dataset
varimax_output_df_varimax_OK2 <- scale(elemzes_input_output_cleaned[,c(27:32, 34:36)]) %*% varimax_rotation_varimax_OK2$loadings

```

```{r}
# Append the principal component scores to the original dataset
elemzes_pca_output_varimax_OK2 <- cbind(elemzes_input_output_cleaned, varimax_output_df_varimax_OK2)

```

Results

```{r}
Korr_output_varimax_OK2 <- cor(
  elemzes_pca_output_varimax_OK2[, c(27:32, 34:36, 38:41)],
  use = "complete.obs"
)

# Visualise the correlation matrix after varimax rotation
library(corrplot)
corrplot(Korr_output_varimax_OK2, method = "number")


```
Export

```{r}
# PNG mentése 300 DPI-ben
png("Korr_output_varimax_OK2.png", 
    width = 22,           
    height = 20, 
    units = "cm", 
    res = 300)

png("Korr_output_varimax_OK2.tiff", 
    width = 22,           
    height = 20, 
    units = "cm", 
    res = 300)

library(corrplot)
corrplot(Korr_output_varimax_OK2, 
         method = "number", 
         number.cex = 0.7,    
         tl.cex = 0.8,        
         tl.col = "black",    
         mar = c(0, 0, 1, 0)) 

dev.off()
```
Summary

```{r}
summary(fokomp_output_varimax_OK2)
```
Examine the principal components using histograms

```{r}
hist(elemzes_pca_output_varimax_OK2$PC1)
hist(elemzes_pca_output_varimax_OK2$PC2)
hist(elemzes_pca_output_varimax_OK2$PC3)
hist(elemzes_pca_output_varimax_OK2$PC4)
```
# LIBRARY DEMAND INDEX AND RANKING

## Preparations

```{r}
# Reverse the sign of PC3
elemzes_pca_output_varimax_OK2 <- elemzes_pca_output_varimax_OK2 %>%
  mutate(PC3 = PC3 * -1)
```
Create histogram

```{r}
hist(elemzes_pca_output_varimax_OK2$PC3)

```
## Construct the Library Demand Index and ranking

```{r}
# 1. Standardisation
# Standardise OK2 and all principal components (PC1, PC2, PC3 and PC4)
elemzes_pca_output_varimax_index <- elemzes_pca_output_varimax_OK2 %>% 
  mutate(
    PC1_std = scale(PC1),
    PC2_std = scale(PC2),
    PC3_std = scale(PC3),
    PC4_std = scale(PC4),
    OK2_std = scale(OK2)
  )

# 2. Unweighted mean
# Calculate the unweighted mean of all standardised components
elemzes_pca_output_varimax_index <- elemzes_pca_output_varimax_index %>%
  mutate(KKI = (PC1_std + PC2_std + PC3_std + PC4_std + OK2_std)/5)

# 3. Ranking
# Rank libraries based on the Library Demand Index
elemzes_pca_output_varimax_index <- elemzes_pca_output_varimax_index %>%
  arrange(desc(KKI)) %>%
  mutate(KKI_rank = min_rank(desc(KKI)))

# Apply min–max scaling to the Library Demand Index (0–100)
elemzes_pca_output_varimax_index <- elemzes_pca_output_varimax_index %>%
  mutate(KKI_100 = (KKI - min(KKI)) / (max(KKI) - min(KKI)) * 100)

# Export the results
write_xlsx(elemzes_pca_output_varimax_index, "elemzes_pca_output_varimax_index.xlsx")

```

## Example analysis

1. Scatter plots illustrating the relationships between principal components

```{r}
KKI100_PC1_OK2 <- ggplot(elemzes_pca_output_varimax_index, aes(x = PC1_std, y = OK2_std)) +
  geom_point(aes(color = KKI_100), size = 2.5, alpha = 0.7) +
  scale_color_gradient(low = "blue", high = "red") +
  
  geom_hline(yintercept = 0, linewidth = 1) +
  geom_vline(xintercept = 0, linewidth = 1) +
  
  coord_fixed() + 
  
  labs(
    x = "PC1: Size of the personal user base",
    y = "OK2: Remote use activity",
    color = "Library Demand Index"
  ) +
  theme_minimal() +
  theme(panel.grid.minor = element_blank())

print(KKI100_PC1_OK2)


ggsave("KKI100_PC1_OK2_matrix.png", 
       plot = KKI100_PC1_OK2, 
       width = 10, 
       height = 8, 
       units = "in", 
       dpi = 300,
       bg = "white") 

ggsave("KKI100_PC1_OK2_matrix.tiff", 
       plot = KKI100_PC1_OK2, 
       width = 10, 
       height = 8, 
       units = "in", 
       dpi = 300,
       bg = "white") 

```

```{r}
KKI100_PC2_PC3 <- ggplot(elemzes_pca_output_varimax_index, aes(x = PC2_std, y = PC3_std)) +
  geom_point(aes(color = KKI_100), size = 2.5, alpha = 0.7) +
  scale_color_gradient(low = "blue", high = "red") +
  
  # 4 részre bontás
  geom_hline(yintercept = 0, linewidth = 1) +
  geom_vline(xintercept = 0, linewidth = 1) +
  
  coord_fixed() + 
  
  labs(
    x = "PC2: Lending activity",
    y = "PC3: Demand for events",
    color = "Library Demand Index"
  ) +
  theme_minimal()

print(KKI100_PC2_PC3)

ggsave("KKI100_PC2_PC3_matrix.png", 
       plot = KKI100_PC2_PC3, 
       width = 10, 
       height = 8, 
       units = "in", 
       dpi = 300,
       bg = "white") 

ggsave("KKI100_PC2_PC3_matrix.tiff", 
       plot = KKI100_PC2_PC3, 
       width = 10, 
       height = 8, 
       units = "in", 
       dpi = 300,
       bg = "white") 
```

# MULTIPLE LINEAR REGRESSION

## Review correlations

Correlation matrix of the input variables and the Library Demand Index (KKI_100)

```{r}
Korr_input_output <- cor(elemzes_pca_output_varimax_index[,c(10:26, 49)], use = "complete.obs")

library(corrplot)
corrplot(Korr_input_output, method = "number")

print(Korr_input_output)

```
## Perform multiple linear regression

Introduce the `TelepulesIndex` variable into the analysis.

```{r}
# Import the settlement development index dataset
telepulesindex <- read_excel("telepulesindex.xlsx")

# Merge the datasets using the AAC10 variable
elemzes_pca_output_varimax_index_ti <- elemzes_pca_output_varimax_index %>%
  left_join(telepulesindex, by = "AAC10")

# Export the merged dataset to an Excel file
write_xlsx(elemzes_pca_output_varimax_index_ti, "elemzes_pca_output_varimax_index_ti.xlsx")

# Inspect the merged dataset
head(elemzes_pca_output_varimax_index_ti)

```
Visualizing the relationship between settlement index (TelepulesIndex) and Library Demand Index

```{r}
library(ggplot2)

KKI100TI <- ggplot(elemzes_pca_output_varimax_index_ti, aes(x = TelepulesIndex, y = KKI_100)) +
  geom_point(color = "steelblue", size = 2.5, alpha = 0.6) +
  

  geom_smooth(method = "lm", color = "black", linetype = "dashed", se = TRUE, linewidth = 0.8) +
  
  labs(
    title = "Settlement development and library demand",
    x = "TelepülésIndex",
    y = "Library Demand Index"
  ) +
  theme_minimal() +
  theme(panel.grid.minor = element_blank())

print(KKI100TI)
```

Calculate the correlation matrix for the variables included in the multiple regression analysis

```{r}
# Calculate the correlation matrix (columns 10–26, 49 and 50)
Korr_input_output_2 <- cor(elemzes_pca_output_varimax_index_ti[, c(10:26, 49, 50)], use = "complete.obs")

library(corrplot)
corrplot(Korr_input_output_2, method = "number", number.cex = 0.7)

print(Korr_input_output_2)

```

### Fit the initial model

Initial model: KKI_reg

```{r}
KKI_reg <- lm(KKI_100 ~ IP1 + IP2 + IP3 + IL1 + IL2 + IK1 + IK2 + IK3 + ISZ1 + ISZ2 + ISZ3 + ISZ4 + ISZ5 + II1 + II2 + IH1 + IR1 + log(TelepulesIndex), data = elemzes_pca_output_varimax_index_ti)
summary(KKI_reg)

#library(lmtest)

# Perform the Breusch–Pagan test
bptest(KKI_reg, studentize = TRUE)
```

Calculate variance inflation factors (VIF)

```{r}
#library(car)
vif(KKI_reg)
```
Since both multicollinearity and heteroskedasticity are present, heteroskedasticity is addressed first using coefficient estimation based on the Generalized Least Squares (GLS) approach.

### GLS

```{r}
# Original regression model
KKI_reg <- lm(KKI_100 ~ IP1 + IP2 + IP3 + IL1 + IL2 + IK1 + IK2 + IK3 + 
                ISZ1 + ISZ2 + ISZ3 + ISZ4 + ISZ5 + II1 + II2 + IH1 + IR1 + 
                log(TelepulesIndex), data = elemzes_pca_output_varimax_index_ti)

elemzes_pca_output_varimax_index_ti$HibatagNegyzet <- residuals(KKI_reg)^2

tabla_white_teszt <- elemzes_pca_output_varimax_index_ti[, c("IP1", "IP2", "IP3", "IL1", "IL2", 
                                                             "IK1", "IK2", "IK3", "ISZ1", "ISZ2", 
                                                             "ISZ3", "ISZ4", "ISZ5", "II1", "II2", 
                                                             "IH1", "IR1", "TelepulesIndex")]

tabla_white_teszt$HibatagNegyzet <- elemzes_pca_output_varimax_index_ti$HibatagNegyzet

negyzetek <- tabla_white_teszt[, 1:(ncol(tabla_white_teszt) - 1)]^2
colnames(negyzetek) <- paste0(colnames(tabla_white_teszt)[1:(ncol(tabla_white_teszt) - 1)], "_negyzete")
tabla_white_teszt <- cbind(tabla_white_teszt, negyzetek)

seged_reg_gls <- lm(log(HibatagNegyzet) ~ ., data = tabla_white_teszt)

omega_elemek <- exp(fitted(seged_reg_gls))

# Fit the weighted regression model
KKI_reg_gls <- lm(KKI_100 ~ IP1 + IP2 + IP3 + IL1 + IL2 + IK1 + IK2 + IK3 + 
                       ISZ1 + ISZ2 + ISZ3 + ISZ4 + ISZ5 + II1 + II2 + IH1 + IR1 + 
                       log(TelepulesIndex), weights = 1/omega_elemek, 
                     data = elemzes_pca_output_varimax_index_ti)

summary(KKI_reg_gls)

```

```{r}
# Perform the Breusch–Pagan test
bptest(KKI_reg_gls, studentize = TRUE)
```
### Stepwise variable selection

```{r}
step_model <- stepAIC(KKI_reg_gls, direction = "both")

summary(step_model)
```

Save the results

```{r}
#library(MASS)

get_significant_vars <- function(model) {
  coef_table <- summary(model)$coefficients
  significant_vars <- rownames(coef_table)[coef_table[, 4] < 0.05]  
  significant_count <- length(significant_vars)                     
  list(names = significant_vars, count = significant_count)
}

initial_summary <- summary(KKI_reg_gls)
initial_aic <- AIC(KKI_reg_gls)
initial_bic <- BIC(KKI_reg_gls)
initial_r2 <- initial_summary$r.squared
initial_adj_r2 <- initial_summary$adj.r.squared
initial_df <- length(KKI_reg_gls$coefficients)
initial_significant <- get_significant_vars(KKI_reg_gls)  
initial_total_vars <- length(initial_summary$coefficients[, 1])  

final_summary <- summary(step_model)
final_aic <- AIC(step_model)
final_bic <- BIC(step_model)
final_r2 <- final_summary$r.squared
final_adj_r2 <- final_summary$adj.r.squared
final_df <- length(step_model$coefficients)
final_significant <- get_significant_vars(step_model)  
final_total_vars <- length(final_summary$coefficients[, 1])  

model_comparison <- data.frame(
  Mutató = c("AIC", "BIC", "R²", "Adjusted R²", "Degrees of freedom",
                "Number of significant variables",
                "Significant variables"),
  `Kezdeti modell` = c(
    initial_aic, initial_bic, initial_r2, initial_adj_r2, initial_df,
    paste(initial_significant$count, "/", initial_total_vars),
    paste(initial_significant$names, collapse = ", ")
  ),
  `Végső modell` = c(
    final_aic, final_bic, final_r2, final_adj_r2, final_df,
    paste(final_significant$count, "/", final_total_vars),
    paste(final_significant$names, collapse = ", ")
  )
)

transposed_table <- t(model_comparison)

write_xlsx(as.data.frame(transposed_table), path = "transzponalt_modell_osszehasonlitas.xlsx")

print(model_comparison)

```

## Model building

```{r}
KKI_reg0 <- lm(KKI_100 ~ IP2 + IP3 + IK1 + IK3 + ISZ1 + ISZ5 + 
    II1 + IH1 + IR1 + log(TelepulesIndex), data = elemzes_pca_output_varimax_index_ti, 
    weights = 1/omega_elemek)
summary(KKI_reg0)
```
Calculate variance inflation factors (VIF)

```{r}
vif(KKI_reg0)
```
Replace IP2 with IP1

```{r}
KKI_reg1 <- lm(KKI_100 ~ IP1 + IP3 + IK1 + IK3 + ISZ1 + ISZ5 + 
    II1 + IH1 + IR1 + log(TelepulesIndex), data = elemzes_pca_output_varimax_index_ti, 
    weights = 1/omega_elemek)
summary(KKI_reg1)

```
Calculate variance inflation factors (VIF)

```{r}
vif(KKI_reg1)
```
Multicollinearity has not decreased.


Remove II1 because p-value > 0.05

```{r}
KKI_reg2 <- lm(KKI_100 ~ IP1 + IP3 + IK1 + IK3 + ISZ1 + ISZ5 + 
    IH1 + IR1 + log(TelepulesIndex), data = elemzes_pca_output_varimax_index_ti, 
    weights = 1/omega_elemek)
summary(KKI_reg2)
```
IP2 + IP3

```{r}
KKI_reg3 <- lm(KKI_100 ~ IP1 + IP2 + IK1 + IK3 + ISZ1 + ISZ5 + 
    IH1 + IR1 + log(TelepulesIndex), data = elemzes_pca_output_varimax_index_ti, 
    weights = 1/omega_elemek)
summary(KKI_reg3)
```
Retain only IP3

```{r}
KKI_reg4 <- lm(KKI_100 ~ IP3 + IK1 + IK3 + ISZ1 + ISZ5 + 
    IH1 + IR1 + log(TelepulesIndex), data = elemzes_pca_output_varimax_index_ti, 
    weights = 1/omega_elemek)
summary(KKI_reg4)
```
Remove ISZ5

```{r}
KKI_reg5 <- lm(KKI_100 ~ IP3 + IK1 + IK3 + ISZ1 + 
    IH1 + IR1 + log(TelepulesIndex), data = elemzes_pca_output_varimax_index_ti, 
    weights = 1/omega_elemek)
summary(KKI_reg5)
```
Calculate variance inflation factors (VIF)

```{r}
vif(KKI_reg5)
```


```{r}
KKI_reg6 <- lm(KKI_100 ~ IP3 + ISZ1 + IK3 +
    IH1 + IR1 + log(TelepulesIndex), data = elemzes_pca_output_varimax_index_ti, 
    weights = 1/omega_elemek)
summary(KKI_reg6)
```
Calculate variance inflation factors (VIF)

```{r}
vif(KKI_reg6)
```
VIF values have decreased.

```{r}
KKI_reg7 <- lm(KKI_100 ~ IP1 + IK3 + ISZ1 + 
    IH1 + IR1 + log(TelepulesIndex), data = elemzes_pca_output_varimax_index_ti, 
    weights = 1/omega_elemek)
summary(KKI_reg7)
```
Calculate variance inflation factors (VIF)

```{r}
vif(KKI_reg7)
```
The VIF values are acceptable. Multicollinearity is no longer present.

RESET test for both models

```{r}
resettest(KKI_reg6)
resettest(KKI_reg7)
```
Breusch-Pagan test for both models

```{r}
bptest(KKI_reg6, studentize = TRUE)
bptest(KKI_reg7, studentize = TRUE)
```
## Model comparison

```{r}
AIC(KKI_reg6, KKI_reg7)
BIC(KKI_reg6, KKI_reg7)
```
AIC: reg6
BIC: reg6

Save the results of the final selected model (REG6)

```{r}
BetaTabla_reg6 <- summary(KKI_reg6)$coefficients

# Sort the table by p-values (column 4)
BetaTabla_reg6[order(BetaTabla_reg6[,4]),]


```
### Model diagnostics

```{r}
#install.packages("performance")
library(performance)
check_model(KKI_reg6, check = "all")
png("KKI_reg6_Model_diagnostics.png",
    width = 3000,
    height = 2400,
    res = 300)
check_model(KKI_reg6, check = "all")
dev.off()

tiff("KKI_reg6_Model_diagnostics.tiff",
     width = 3000,
     height = 2400,
     res = 300,
     compression = "lzw")
check_model(KKI_reg6, check = "all")
dev.off()
```

```{r}
residuals <- residuals(KKI_reg6)

# QQ plot for assessing the normality of the residuals
qqnorm(residuals, main = "QQ Plot of Residuals")
qqline(residuals, col = "red", lwd = 2)

# Histogram
hist(residuals, breaks = 30, main = "Histogram of Residuals", xlab = "Residuals", col = "lightblue", border = "black")

# Shapiro–Wilk test for normality
shapiro_test <- shapiro.test(residuals)
print("Shapiro-Wilk test results:")
print(shapiro_test)

# Kolmogorov–Smirnov test for normality
ks_test <- ks.test(residuals, "pnorm", mean = mean(residuals), sd = sd(residuals))
print("Kolmogorov-Smirnov test results:")
print(ks_test)

# Diagnostic visualisation
par(mfrow = c(1, 2))  
qqnorm(residuals, main = "QQ-Plot")
qqline(residuals, col = "red")
hist(residuals, breaks = 30, main = "Histogram", xlab = "Residuals", col = "lightblue", border = "black")

```

Export the REG6 model results to an Excel file

```{r}
reg6_modell_eredmenyek <- as.data.frame(summary(KKI_reg6)$coefficients)

write_xlsx(reg6_modell_eredmenyek, "reg6_modell_eredmenyek.xlsx")
```



